Title of the report

### 2011-12-02 doc/report/report.tex
report!

\begin{document}
% ska gå att förstå och locka till läsning
%ska tydligt beskriva vad rapporten handlar om
%ska innehålla vettiga sökord
\title{Scaling up a GF grammar and preparing it for cool stuff for Swedish}
\author{Malin Ahlberg}
\maketitle
\newpage

@@

\subsection{Saldo}
A good parser needs a good lexicon. We have used of Saldo\cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built upon Svenskt Associations Lexicon and contains information about more than 
120 000 modern Swedish words.
The words are equipped with both sematic and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of sematically connected words.\\
\newpage % only until saldo can be on same page as the pic
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}
For our purpose however, only the morphological data was needed.
Saldo provides this in XML format under LPGL. The data can be processed and
translated to GF format as described in 
section \ref{sec:prog.saldo}.
%Other usages?

@@

\subsubsection{The resource library}
The resource library\cite{gf-resource} is a very useful part of the framework. It is provided in the 
GF package. %, they give a good start for all who want to write an application grammar.
Instead of letting the users of the framework reimplement the morphological and
the fundamental syntactic structures of a language for each application, all
this can be found as libraries. So far there are 20 languages in the library, and
they all share the same abstract syntax.
The resource grammars describe how to construct phrases and sentences and how to
decline words. The latter is done by smart paradigms: functions that analyses
some given forms of a word to find out how the inflection table should look.
For example, the declination of many Swedish nouns can be determined by looking
at the singular form only. This is the case for ``flicka"??, the correct declination
can be guessed by noting that the word ends with an ``a". Like most Swedish nouns
ending with an ``a", it follows the first?? declination. For others, like
"hjärta", also the plural form "hjärtan" is needed, even though it ends
with an ``a". The worst case is nouns that need four forms, both singular and
plural in definite and indefinite form.
There are also a small test lexicon included, containing some hundered of common words.\\

As an open source-project, GF is constantely being developed and improved. New
languges are added, the compiler is being improved, ways of använad det på ett
lätt sätt are added and the possiblities to use GF in different environments
increased. There are  research about how to make better use of the dependent
types, for solving syllogism or generating natural language via Montague
semantics.
Further, experiments on using GF for free parsing has been conducted \cite[\textsection 4.1]{gfMech} 
This is also a future goal for our project and one of the delmål has been to
model an important part of Swedish in GF. The starting point was the resource
grammar and the result should still have a strong connection to this.

@@

\subsection{Talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being an freely avaible, manually annotade, large scale treebank,
Talbanken\cite{talbanken} was perfect for this ändamål.
It was created in the 1970s at Lund University. In total it consists of four
parts. Two of them are transcribitions of spoken language, one a collection of
text written by high school students, and the last one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.\\
Talbanken was also used to train the Swedish version of the Maltparser\cite{malt}
and was thus redistributed in a more modern version\cite{talbanken05} This work
was done by Nivre, Hall and Nilsson in 2005. They released it in Malt and Tiger XML-formats,
where the trees had been made deeper and more detailed. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout this
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

@@

The basic algorithm for porting Saldo is the same as Angelov's. It makes use of
the smart paradigms in GF : explanation.
%Find the paper!
%To understand it, we first need to look at differences between Saldo and a GF lexicon.
The program will try to feed as few forms as possible to the GF smart paradigm.
It starts from the easiest case, compare the results and either saves the code or 
makes another try using more word forms.
Figure \ref{pic:TabVax} shows a the information given by Saldo and by GF respectively.\\
\begin{figure}[h]
  \begin{center}
\subfloat[Saldo]{\label{pic:saldoVax}\includegraphics[height=90mm]{saldoVax.png}}
\hspace{5mm}
\subfloat[GF]{\label{pic:gfVax}\includegraphics[height=60mm]{gfVax.png}}
\caption{Inflection table for the verb ``växa" (grow)}
\label{pic:TabVax}
  \end{center}
\end{figure}
We can see that there are more saldo forms and some are only in GF.

Explanation of the algorithm: map categories, try each paradigm, compare the table of forms
created. If several forms in saldo, gf should pick on of them. Saldo has forms for
compounds, gf does not. A grammar is written, the correct ones saved, the others 
retried. Try all paradigms which can be formed from saldo. 
There are a number of differences between and saldo Till skillnad från GF,
saldo does not contain forms that aren't used gf generates all forms. A form
needed to generate the gf table may not be given in saldo such as singular
forms for 'byxor'. Also contains idioms, 'hålla huvudet kallt' and has
different grammatical analyse, a lot of pronouns. Example.

In Saldo there is also semantic info, so
We need the numbers from saldo if there are to lemmas with the same
identifier. växa (bli större), växa (vara en planta) vs. sluta\_V and sluta2\_V
(slöt). However gf does not want more than one table
if the forms are identical, saldo's morpho neither, but there are five in big saldo. 
We are just interested in morphology, so far. Dont want too big lexicon.

The new version have also been made more robust, print logs and list of all
words that wasn't imported together with a reason. Reson could be that it
cannot construct all forms or that the word class shouldn't be in the gf lexicon.
Costs very much memory, the saldo file is big, the generated lexicon is big.
Saldo is divided to save memory. Errors should just be reported, but if something oförutsett
happens and one part fail, the separation ensures that the others wont, also
possible to restart. Problems with naming convention, special
characters (å,ä,ö,apostrop,entrecôte.) needs to be translated , but avoid
classhes 'kaeltisk' 'kältisk', gf will crash by this, so we add a 1
in the end whenever we have changed a letter.

Other word classes we do not import and why.
Those should not be in the standard gf lexicon but maybe somewhere else.
The list can be used for adding other kinds of lexicons - idioms, have pos-tags.
Also to analyse what's wrong, or use the words with a word guesser. The code
has been adapted for this.
For exapmle pronouns - dets, have combined information from Talbanken to add them in the correct
category. How this works, look at tags. Small amount, can be done manually.

The saddest thing: no valency info. Crucial for GF. Only reflexive and verbs with particles
have any information. The earlier explained techniques may be used for this in Talbanken.
To get even bigger material, use Korp, but then no guarantee.

@@

\subsection{Results}
How many words, how many Talbanken words are not here now? Why?

\section{The grammar}
\label{sec:prog.grammar}
Although we can't expect to get a full coverage, we would like a tillfredsställande
Explain what's in the resources, constructions present in most languages. 
Been working on a grammar fragment based on Talbanken,
like cool people like Montague. May also be interesting for other purposes.
Swedish shares with Nor and Dan. Extra module, covering things like sw is a
v2-lang, focusing parts of sentences.

\subsection{The Swedish resource grammar}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology and many phrase and clause constructions. Most of the implementation
is shared with the other scandinavian languages.
Apart form descriping morphology, the resource grammar deals with
word order, agreement, tempus, basic conjunction etc.
\enumsentence{\shortexnt{8}
{Har& han& inte& ätit& de & gula& äpplena& idag?}
{Has& he& not& eaten& the& yellow& apples& today?}} %correct!!!
\label{gfSwe:parseable}
The grammar can construct and parse the sentence in example \ref{gfSwe:parseable}
and the parse tree is represented as in figure \ref{gfSwe:parsetree}.
%Covers sentences like
% better sentence where the verb phrase is split up?
\begin{figure}[h]
\includegraphics[width=70mm,height=70mm]{apples.pdf}
\caption{Parse tree for ``Har han inte ätit de gula äpplena idag".}%correct
\label{gfSwe:parsetree}
\end{figure}
Even though the verb phrase \emph{``har inte ätit.."} is discontinuous, the whole
phrase is still treated as one constituent in GF. The parts
are connected in the tree, and the subject \emph{``han"} is put between the
finite verb and the rest of the phrase. At the code level, this is implemented
using records.
The linearization rule for questions -- clauses with inverted order -- picks 
the field for the finite verb and appends it to the subject followed by
the negation, infinite part and the complement of the verb phrase: \\
\verb|table { Inv => verb.fin ++ subj ++ verb.neg ++ verb.inf ++ verb.compl ; ... | \\

@@

\subsection{Added}
Summary of what have been added and motivation.

@@

\subsubsection{Quantifiers, determiners and predeterminers}
The GF analyse differs sometimes. Experiments of how to extract
this information from Talbanken by using Saldo as a starting poin.
One of the biggest differences between the GF analysis and the one made in Talbanken
is related to pronouns. While GF only recognizes the personal pronouns, 
`sådan', `fler`, `somliga' etc. are all regarded as pronouns in Talbanken and Saldo.
In the GF analyse they may be quantifiers, determiners or predeterminers. 
All pronouns from Saldo was listed and by searching Talbanken for which tags
they got we could sort out the ones acting either as
determiners, as `sådana' in sentence \ref{ex:QuantDet}, or as adjectives as
`sådan' in sentence \ref{ex:QuantAdj}. 
\enumsentence{Sådana katter vill jag ha}\label{ex:QuantDet}
\vspace{-3mm}
\enumsentence{Han är sådan}\label{ex:QuantAdj}
\vspace{-3mm}
\enumsentence{Sådant är inte bra}\label{ex:QuantDet2}
Since all determiners can be used alone as noun phrases, sentence
\ref{ex:QuantDet2} is also accepted.
%what is the difference really???
%The techniques of searching Talbanken can be used
%for this purpose as well. 
%combining the information in Saldo and Talbanken
%pronouns/Quant. Many, created categories for 'sådan', and 'fler'. Not many in each,
%has to decide, one category for each?

@@

\chapter{Summary}
\section{Results and discussion}
Results from mapping, can probably be improved, bit by bit, depends much on words.
Valencys would therefor help.  
Saldo, made a big difference to renew the lexicon. Tested what was missing now from talbanken,
results. Mostly compound nouns, which we can't expect in the dictionary.
Is too slow, can't be used for parsing right away

Grammar - hard to evaluate automatically, but Elibet is an expert who has been involved
in the process to verify the solutions. Testing trees against talbanken.
The big lexicon makes it very slow, eats all memory

Discussion of the results and methods, why are the results good, why are they not better.
What other methods could have been used? What did we/I expect and what happend?

\section{Future Work}
\label{sec:future}
What is still left to do? Description and direction of future work:
Handling unknown words and , handling unknown grammar, 
how to make the parser robust, elipses and long sentences, names and numbers.
What is still needed in the grammar, how important that is. 
\subsection{Lexicon/Parallell parsing}
Speed up.
\subsection{Valency}
\label{sec:futureValency}
how to get valency information 'sitter och läser',
is this doable for swedish?
description of extracting this from talbanken/korp
or handling it by the robust parser or the grammar itself (leaveout obj?).
make the parser faster by using smaller lexicon.
Lexicon tool on the fly.
\subsection{Probabilites}
\label{sec:futureProbabilities}
Getting and using probabilities, to disambiguate.
process list from saldo.
Use multiligual treebank?
stop ambiguities by fixing so that Gen of a quant isn't a quant, or better, that all fields
have a genitive field.
More fields in the grammar? Or leave it all to the robust parser

\section{Conclusion}
There is obviously a lot do still, will never get finish with NPL.
Grammar may need even bigger changes, to allow enough but not too much.
Different from writing a grammar for generation.
Results from mapping shows that the translation is doable, but becomes harder
since talbanken don't have formal rules.
The saldo shows how easy to make use of utomstående resources, promising if
we want to extract other information. 
Hard, but not finished!

>>>>

\begin{document}
% ska gå att förstå och locka till läsning
%ska tydligt beskriva vad rapporten handlar om
%ska innehålla vettiga sökord
\title{Title of the report} %A nice titleScaling up a GF grammar and preparing it for cool stuff for Swedish}
\author{Malin Ahlberg}
\maketitle
\newpage

@@

\subsection{Saldo}
A good parser needs a good lexicon. We have used of Saldo\cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built upon Svenskt Associations Lexicon and contains information about more than 
120 000 modern Swedish words.
The words are equipped with both sematic and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of sematically connected words.\\
\newpage % only until saldo can be on same page as the pic
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

If one word
have different meanings, there will be separate entries for each of them.
The word ``växa" (``grow") for example, has two meanings: \emph{getting bigger}
\enumsentence{Gräset växte varje dag\\
The grass grew each day}
or \emph{being a plant}. 
\enumsentence{Det växte en gammal ek där.\\
There stood an old oak.}
The word ``växa" is however declined the same way for both interpretations of the word.
Hence Saldo's morpological lexicon, provided under LPGL in XML format,
only keeps one entry of ``växa".
For the homonyms that does not share the whole inflection tables, different entries are 
kept in the morphological lexicon.

For our purpose, only the morphological lexcion was needed.
The data can be processed and translated to GF format as described in 
section \ref{sec:prog.saldo}.

@@

\subsubsection{The resource library}
The resource library\cite{gf-resource} is a very useful part of the framework. It is provided in the 
GF package. %, they give a good start for all who want to write an application grammar.
Instead of letting the users of the framework reimplement the morphological and
the fundamental syntactic structures of a language for each application, all
this can be found as libraries. So far there are 20 languages in the library, and
they all share the same abstract syntax.
There are also a small test lexicon included, containing some hundered of common words.

The resource grammars describe how to construct phrases and sentences and how to
decline words. The latter is done by smart paradigms: functions that analyses
some given forms of a word to find out how the inflection table should look.
For example, the declination of many Swedish nouns can be determined by looking
at the singular form only. 
\begin{wrapfigure}{l}{0.4\textwidth}
\begin{tabular}{| l | l |}
\hline
1st declination & 5th declination \\
\hline
flick\textbf{a}    &     hjärt\textbf{a}   \\
flick\textbf{an}    &    hjärt\textbf{at}  \\
flic\textbf{kor}    &    hjärt\textbf{an}  \\
flic\textbf{korna}  &    hjärt\textbf{ana} \\
\hline
\end{tabular}
\end{wrapfigure}
This is the case for ``flicka", the correct declination
can be guessed by using the first declination. % and noting that the word ends with an ``a". 
%Like most Swedish nouns ending with an ``a", it follows the first declination.
For others, like "hjärta", also the plural form "hjärtan" is needed.
%, even though it ends with an ``a". 
The worst case is nouns that need four forms, both singular and
plural in definite and indefinite form.

\vspace{5mm}
As an open source-project, GF is constantely being developed and improved. New
languges are added, the compiler is being improved, ways of använad det på ett
lätt sätt are added and the possiblities to use GF in different environments
increased. There are  research about how to make better use of the dependent
types, for solving syllogism or generating natural language via Montague
semantics.
Further, experiments on using GF for free parsing has been conducted \cite[\textsection 4.1]{gfMech} 
This is also a future goal for our project and one of the delmål has been to
model an important part of Swedish in GF. The starting point was the resource
grammar and the result should still have a strong connection to this.

@@

\subsection{Talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being an freely avaible, manually annotade, large scale treebank,
Talbanken\cite{talbanken} was perfect for our purpose.
It was created in the 1970s at Lund University. In total it consists of four
parts. Two of them are transcribitions of spoken language, one a collection of
text written by high school students, and the last one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.\\
Talbanken was also used to train the Swedish version of the Maltparser\cite{malt}
and was thus redistributed in a more modern version\cite{talbanken05} This work
was done by Nivre, Hall and Nilsson in 2005. They released it in Malt and Tiger XML-formats,
where the trees had been made deeper and more detailed. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout this
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

@@

The basic algorithm for porting Saldo is the same as Angelov's, it works by
making use of the smart paradigms in GF.
%Find the paper!
%To understand it, we first need to look at differences between Saldo and a GF lexicon.
The program will try to create a lexicon by finding a number of forms for every
word and feeding them to the smart paradigms. As few forms as possible should be
used, and the resulting table generated by GF should be equivalent to the one found
in Saldo. This process is done iteratively,
starting from the easiest case. For each attempt the results are compared and when
the correct way of using a word has been found, the code is saved.
Figure \ref{pic:TabVax} shows a the information given by Saldo and by GF respectively.\\
\begin{figure}[h]
  \begin{center}
\subfloat[Saldo]{\label{pic:saldoVax}\includegraphics[height=90mm]{saldoVax.png}}
\hspace{5mm}
\subfloat[GF]{\label{pic:gfVax}\includegraphics[height=60mm]{gfVax.png}}
\caption{Inflection table for the verb ``växa" (grow)}
\label{pic:TabVax}
  \end{center}
\end{figure}
The tables does not overlap completely, there are some more forms in Saldo's table (eg. the 
compounding form ``väx-")
while the one generated in GF contains some that are not in Saldo (eg. ``vuxits").
Saldo may also contain variats: ``växt" and ``vuxit" are both supine form.
As far as possible, the program makes up for this by only comparing the overlapping forms
and only requiring that GF generatens one variant whenever alternatives are given. 
The program continues to generate new lexicons as long as there are more
paradigms to choose from and compare the results after each run.

The program has been made more robust than the previous version, it prints
logs and finally lists of the words that were not imported. The logs provides
information about which paradigms that were used for each word and the result
from comparing the inflection tables.
During the porting of Saldo, some renaming of the lemmas was needed.
Special characters like \emph{å,ä,ö} should not occure in GF function names.
Those are translated into \emph{aa,ae,oe} respectively. 
The Saldo identifier \verb|äta..vb| is thus turned into the GF identifier \verb|aeta_V|.
Since there might be two lemmas that share the same
identifier after the translation, a number is added to the name whenever a translation
have been done: 
\begin{tabular}{lll}
kältisk & $\rightarrow$ & kaeltisk\_1 \\
kaeltisk & $\rightarrow$ & kaeltisk \\
entrecôte &$\rightarrow$ & entrecoote\_1 \\ 
\end{tabular}

%Problems with naming convention, special
%characters (å,ä,ö,apostrop,entrecôte.) needs to be translated , but avoid
%classhes 'kaeltisk' 'kältisk', gf will crash by this, so we add a 1
%in the end whenever we have changed a letter.
%,é
%Costs very much memory, the saldo file is big, the generated lexicon is big.
%Saldo is divided to save memory. Errors should just be reported, but if something oförutsett
%happens and one part fail, the separation ensures that the others wont, also
%possible to restart.

%Explanation of the algorithm: map categories, try each paradigm, compare the table of forms
%created. If several forms in saldo, gf should pick on of them. Saldo has forms for
%compounds, gf does not. A grammar is written, the correct ones saved, the others 
%retried. Try all paradigms which can be formed from saldo. 
%In Saldo there is also semantic info, so
%We need the numbers from saldo if there are to lemmas with the same
%identifier. växa (bli större), växa (vara en planta) vs. sluta\_V and sluta2\_V
%(slöt). However gf does not want more than one table
%if the forms are identical, saldo's morpho neither, but there are five in big saldo. 
%We are just interested in morphology, so far. Dont want too big lexicon.

@@

\subsection{Results}
All words from Saldo were not be imported. There are a number of reasons
for this, the most obvious one is that we do not want all categories from
Saldo in the GF lexicon. Prepositions(\emph{pp}) and numerals(\emph{nl}) etc. are
assumed to be known already,
and are usually not kept in the standard lexicon, but in modules for the basic
structures of the language. Categories involving several words(\emph{vbm,nnm})
need to be handeld differently, usually as idioms. Pronouns(\emph{pn}) are not  
analysed the same way, see section \ref{sec:swegf}. Some experiments on finding
the right GF-category for those have been done, see \ref{sec:gf.quant}.
Six categories were imported: \\
\begin{tabular}{l|lll}
& Saldo tag & GF category & Example \\
\hline
 Adverb & \textbf{ab} &\textbf{Adv} & ofta \\
 Adjective&\textbf{av} &    \textbf{A} & gul\\
 Noun & \textbf{nn} &\textbf{N} & hus\\
 Verb & \textbf{vb} &\textbf{V} & springa\\
 Reflexive verbs  &\textbf{vbm}& \textbf{V2} & raka sig\\
 Verbs with particles &\textbf{vbm}& \textbf{V}  & hitta på\\
\end{tabular}\\

%
%There are a number of differences between and saldo Till skillnad från GF,
%saldo does not contain forms that aren't used gf generates all forms. A form
%needed to generate the gf table may not be given in saldo such as singular
%forms for 'byxor'. Also contains idioms, 'hålla huvudet kallt' and has
%different grammatical analyse, a lot of pronouns. Example. 
One reason for why words of the categories listed above could not be imported
is that Saldo does not contain information about word forms that are not used.
In some cases those forms are needed by the smart paradigm. 
This was a common reason why nouns could not be imported. 
Consider for example
the plural tantum %\cite[44]{SAG}
noun ``glasögon" (``glasses"). Since the smart paradigm
requires a singular form, this lemma could not be added to the lexicon. ``Glasögon" could
be added manually, by using the somewhat incorrect singular form ``glasöga" (``glass-eye").
The same problem occured for verbs that always are used in passive form:
``andas" (``breathe"), ``frodas" (``flourish"), ``lyckas" (``succeeed"). This
was the case for close to 80\% of the failing verbs of type \verb_vb_.

In some cases the smart paradigms could not generate the correct declination,
although this happend seldom. % frysa,fisa passive preterium

The resulting dictionary contains more than 100 000 entries, compared to the total size
of Saldo this is approximately 80\%. When testing the coverage of Talbanken,
we found that 3058 words were still missing. In addition, there were 1817 words
which were given different labels in GF than in Talbanken. Many of those are
acceptable, such as the ones in table \ref{tab:saldodiff} and reflects the difference 
made in the analyses, but others are examples of words that are still missing from
the lexicon.
\begin{figure}[h]
\begin{tabular}{lll}
Word & Talbanken tag & GF category \\
\hline
måste & MVPS          & VV \\
allting & POTP & N \\
få & POZP & Det \\
\end{tabular}
\caption{}
\label{tab:saldodiff}
\end{figure}
Among the words that were completely missing, more than 2000 was found to be nouns,
many of which were compounds.

The lexicon does not contain any valency information since Saldo does not give any such
information. Valencies are however crucial for GF, and by using the results from the mapping
(see section \ref{sec:Mapping}) it should be possible to extract this information about this.
%The saddest thing: no valency info. Crucial for GF. Only reflexive and verbs with particles
%have any information. The earlier explained techniques may be used for this in Talbanken.
%To get even bigger material, use Korp, but then no guarantee.

\section{The grammar}
\label{sec:prog.grammar}
Although we can't expect to get a full coverage, we aim for a high-quality 
grammar with satisfactory coverage.
We have been working on a grammar fragment which have been adapted to cover
constructions used in Talbanken. The starting point has been the Swedish
resource grammar and the new grammar should still be compatible with this.
A description of the Swedish implementation of the resource grammar is given
in the next section, while section \ref{sec:Added} shows and explains the
development during this project.
%Explain what's in the resources, constructions present in most languages. 
%Been working on a grammar fragment based on Talbanken,
%like cool people like Montague. May also be interesting for other purposes.
%Swedish shares with Nor and Dan. Extra module, covering things like sw is a
%v2-lang, focusing parts of sentences.

\subsection{The Swedish resource grammar}
\label{sec:swegf}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology and many phrase and clause constructions. Most of the implementation
is shared with the other scandinavian languages.
Apart form descriping morphology, the resource grammar deals with
word order, agreement, tempus, basic conjunction etc.
\enumsentence{\shortexnt{8}
{Har& han& inte& ätit& de & gula& äpplena& idag?}
{Has& he& not& eaten& the& yellow& apples& today?}} 
\label{gfSwe:parseable}
The grammar can construct and parse the sentence in example \ref{gfSwe:parseable}
and the parse tree is represented as in figure \ref{gfSwe:parsetree}.
%Covers sentences like
% better sentence where the verb phrase is split up?
\begin{figure}[h]
\includegraphics[width=70mm,height=70mm]{apples.pdf}
\caption{Parse tree for ``Har han inte ätit de gula äpplena idag".}
\label{gfSwe:parsetree}
\end{figure}
Even though the verb phrase \emph{``har inte ätit.."} is discontinuous, the whole
phrase is still treated as one constituent in GF. The parts
are connected in the tree, and the subject \emph{``han"} is put between the
finite verb and the rest of the phrase. At the code level, this is implemented
using records.
The linearization rule for questions -- clauses with inverted order -- picks 
the field for the finite verb and appends it to the subject followed by
the negation, infinite part and the complement of the verb phrase: \\
\verb|table { Inv => verb.fin ++ subj ++ verb.neg ++ verb.inf ++ verb.compl ; ... | \\

@@

\subsection{Added}
\label{sec:Added}
Summary of what have been added and motivation.

@@

\subsubsection{Quantifiers, determiners and predeterminers}
\label{sec:gf.quant}
The GF analyse differs sometimes. Experiments of how to extract
this information from Talbanken by using Saldo as a starting poin.
One of the biggest differences between the GF analysis and the one made in Talbanken
is related to pronouns. While GF only recognizes the personal pronouns, 
`sådan', `fler`, `somliga' etc. are all regarded as pronouns in Talbanken and Saldo.
In the GF analyse they may be quantifiers, determiners or predeterminers. 
All pronouns from Saldo was listed and by searching Talbanken for which tags
they got we could sort out the ones acting either as
determiners, as `sådana' in sentence \ref{ex:QuantDet}, or as adjectives as
`sådan' in sentence \ref{ex:QuantAdj}. 
\enumsentence{Sådana katter vill jag ha}\label{ex:QuantDet}
\vspace{-3mm}
\enumsentence{Han är sådan}\label{ex:QuantAdj}
\vspace{-3mm}
\enumsentence{Sådant är inte bra}\label{ex:QuantDet2}
Since all determiners can be used alone as noun phrases, sentence
\ref{ex:QuantDet2} is also accepted.
%what is the difference really???
%The techniques of searching Talbanken can be used
%for this purpose as well. 
%combining the information in Saldo and Talbanken
%pronouns/Quant. Many, created categories for 'sådan', and 'fler'. Not many in each,
%has to decide, one category for each?

@@

\chapter{Summary}
\section{Results and discussion}
%Results from mapping, can probably be improved, bit by bit, depends much on words.
%Valencys would therefor help.  
%Saldo, made a big difference to renew the lexicon. Tested what was missing now from talbanken,
%results. Mostly compound nouns, which we can't expect in the dictionary.
%Is too slow, can't be used for parsing right away
%
%Grammar - hard to evaluate automatically, but Elibet is an expert who has been involved
%in the process to verify the solutions. Testing trees against talbanken.
%The big lexicon makes it very slow, eats all memory
%
%Discussion of the results and methods, why are the results good, why are they not better.
%What other methods could have been used? What did we/I expect and what happend?

\section{Future Work}
\label{sec:future}
What is still left to do? Description and direction of future work:
Handling unknown words and , handling unknown grammar, 
how to make the parser robust, elipses and long sentences, names and numbers.
What is still needed in the grammar, how important that is. 
\subsection{Parallell parsing/Chunk parsing}
\subsection{Lexicon}
%Speed up!
%The list of not-imported can be used for adding other kinds of lexicons - idioms, have pos-tags.
%Also to analyse what's wrong, or use the words with a word guesser. The code
%has been adapted for this.
%For exapmle pronouns - dets, have combined information from Talbanken to add them in the correct
%category. How this works, look at tags. Small amount, can be done manually.

\subsection{Valency}
\label{sec:futureValency}
%how to get valency information 'sitter och läser',
%is this doable for swedish?
%description of extracting this from talbanken/korp
%or handling it by the robust parser or the grammar itself (leaveout obj?).
%make the parser faster by using smaller lexicon.
%Lexicon tool on the fly.

\subsection{Probabilites}
\label{sec:futureProbabilities}
%Getting and using probabilities, to disambiguate.
%process list from saldo.
%Use multiligual treebank?
%stop ambiguities by fixing so that Gen of a quant isn't a quant, or better, that all fields
%have a genitive field.
%More fields in the grammar? Or leave it all to the robust parser

\section{Conclusion}
%There is obviously a lot do still, will never get finish with NPL.
%Grammar may need even bigger changes, to allow enough but not too much.
%Different from writing a grammar for generation.
%Results from mapping shows that the translation is doable, but becomes harder
%since talbanken don't have formal rules.
%The saldo shows how easy to make use of utomstående resources, promising if
%we want to extract other information. 
%Hard, but not finished!
